\clearpage
\addcontentsline{toc}{chapter}{Bibliography}
\begin{thebibliography}{9}
        \footnotesize{
        \bibitem{bib:wikiTimeDomain} Wikipedia, \underline{https://en.wikipedia.org/wiki/Time\_domain}, 2022.
        \bibitem{bib:discreteTimeSystems} Wikipedia, \underline{https://en.wikipedia.org/wiki/Discrete\_time\_and\_continuous\_time}, 2022.
        \bibitem{bib:floatingPointPrecisionIssues} Wikipedia, \underline{https://en.wikipedia.org/wiki/Floating-point\_arithmetic\#Accuracy\_problems}, 2022.
        \bibitem{bib:ltiSystems} Wikipedia, \underline{https://en.wikipedia.org/wiki/Linear\_time-invariant\_system}, 2022.
        \bibitem{bib:frequencyDomain} Wikipedia, \underline{https://en.wikipedia.org/wiki/Frequency\_domain}, 2022.
        \bibitem{bib:fourierTransform} Wikipedia, \underline{https://en.wikipedia.org/wiki/Fourier\_transform}, 2022.
        \bibitem{bib:discreteFourierTransform} Wikipedia, \underline{https://en.wikipedia.org/wiki/Discrete\_Fourier\_transform}, 2022.
        \bibitem{bib:shortTimeFourierTransform} Wikipedia, \underline{https://en.wikipedia.org/wiki/Short-time\_Fourier\_transform}, 2022.
        \bibitem{bib:wernerHeisemberg} Wikipedia, \underline{https://en.wikipedia.org/wiki/Uncertainty\_principle}, 2022.
        \bibitem{bib:discreteCosineTransform} Wikipedia, \underline{https://en.wikipedia.org/wiki/Discrete\_cosine\_transform}, 2022.
        \bibitem{bib:fastHaarTransform} Wikipedia, \underline{G. Kaiser, "The fast Haar transform," in IEEE Potentials,} \underline{vol. 17, no. 2, pp. 34-37, April-May 1998, doi: 10.1109/45.666645.}, 2022.
        \bibitem{bib:zTransform} Wikipedia, \underline{https://en.wikipedia.org/wiki/Z-transform}, 2022.
        \bibitem{bib:zTransformExamples} Richard Baraniuk et al.\ \underline{https://eng.libretexts.org/Bookshelves/Electrical\_Engineering/} \underline{Signal\_Processing\_and\_Modeling/Signals\_and\_Systems\_(Baraniuk\_et\_al.)} \underline{/12\%3A\_Z-Transform\_and\_Discrete\_Time\_System\_Design/12.06\%3A\_Region\_of} \underline{\_Convergence\_for\_the\_Z-Transform}, Rice University, May 23, 2022.
        \bibitem{bib:eigenFunctions} Wikipedia, \underline{https://en.wikipedia.org/wiki/Eigenfunction}, 2022.
        \bibitem{bib:filtering} Wikipedia, \underline{https://en.wikipedia.org/wiki/Filter\_(signal\_processing)}, 2022.
        \bibitem{bib:linearPhase} Wikipedia, \underline{https://en.wikipedia.org/wiki/Linear\_phase}, 2022.
        }
\end{thebibliography}

